Titanic Kaggle Competition
========================================================

The task is to predict whether a given passenger survived the sinking of the Titanic based on available data. The data includes passenger's age, gender, name, number of family members on board (parent or chile and spouse or sibling), the fare that was paid, class of travel, ticket number and cabin (for some), and the port at which they boarded the Titanic.

The data is split into two sets -- train set (891 observations) and test set (418 observations). The solutions are evaluated by comparing the percentage of correct answers against the test subset.

I decided to approach the problem by building a decision tree model.

```{r intro}
library(randomForest)
library(party)
set.seed(415) 
```

Loading train and test data and combining them into one dataset for subsequent clean-up and manipulation.

```{r loadingdata}
train <- read.csv(file="train.csv")
test <- read.csv(file="test.csv")
test$Survived <- NA
combi <- rbind(train, test)
```

Additional engineered attributes were introduced:
- A passenger Title (such as "Miss", "Mrs", "Master", "Mr", "Lady", "Sir", "Dr", etc.) was extracted from the Name attribute.
- FamilyID to link family members together -- the passenger name was extracted Name attribute, the total number of family members was calculated from Parch and SibSp attributes (Parch + SibSp + 1). The idea was that known fate of other family members would be a good predictor of survival likelihood, especially for large families that would have more troubles staying together. There was a large number of small families, so families with small number of members were grouped into Single and TwoMembers categories.

This portion of the script is based on <a href="trevorstephens.com/post/72916401642/titanic-getting-started-with-r">Trevor Stephen's tutorial</a>. One change I made was to merge Mme and Mlle (French titles) with their English counterparts. I also experimented with splitting groups based on FamilyID into Single, Two persons, Three persons groups, but that did not seem to improve the score. 

```{r newvars}
combi$Name <- as.character(combi$Name)
#Title
combi$Title <- sapply(combi$Name, FUN=function(x) {strsplit(x, split='[,.]')[[1]][2]})
combi$Title <- sub(' ', '', combi$Title)
combi$Title[combi$Title %in% c('Miss', 'Mlle')] <- 'Miss' #normalizing French titles - Madmoiselle
combi$Title[combi$Title %in% c('Mme', 'Mrs')] <- 'Mrs' #normalizing French titles - Madame
combi$Title[combi$Title %in% c('Capt', 'Don', 'Major', 'Sir')] <- 'Sir' #combining gentlemen
combi$Title[combi$Title %in% c('Dona', 'Lady', 'the Countess', 'Jonkheer')] <- 'Lady' #combining ladies
combi$Title <- factor(combi$Title)
table(combi$Title)
# FamilyID
combi$FamilySize <- combi$SibSp + combi$Parch + 1 # Calculating Family size 
combi$Surname <- sapply(combi$Name, FUN=function(x) {strsplit(x, split='[,.]')[[1]][1]}) #Extracting family name as surname
combi$FamilyID <- paste(as.character(combi$FamilySize), combi$Surname, sep="") #Pasting surname and family size together for family ID
combi$FamilyID1 <- combi$FamilyID # Creating a copy for manipulation
combi$FamilyID1[combi$FamilySize <= 2] <- 'Small' #Not concerned with small families, combining all under 2 members into one 'Small' group.
famIDs <- data.frame(table(combi$FamilyID)) #Checking frequency of family IDs and whether they match the family size
famIDs <- famIDs[famIDs$Freq <= 2,] #Limiting famID to small families less than people (this introduces noise)
combi$FamilyID1[combi$FamilyID1 %in% famIDs$Var1] <- #Overwriting family names for small families with 'Small' 
combi$FamilyID1 <- factor(combi$FamilyID1)
```

Using new FamilyID1 and FamilySize attributes, the random forest and conditional inference models were run (also, based on Trever's tutorial).

The first random tree attributes calculated Survival likelihood for train data based on gender, age, whether siblings/spouse and parents/children were on board, fare paid, the port where the passenger embarked:

   Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID

The resulting model was then used to predict the outcome for the test data.

```{r randomforest1}
#Random forest tree:
fit <- randomForest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + 
                        Embarked + Title + FamilySize +
                        FamilyID2, data=train, importance=TRUE, ntree=2000)
prediction <- predict(fit, test)
submit <- data.frame(PassengerId = test$PassengerId, Survived = prediction)
write.csv(submit, file = "firstforest_m1.csv", row.names = FALSE)
```


```{r condinftree1}
#Conditional inference tree:
fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + Fare + Title + 
                   FamilySize + FamilyID + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = prediction)
write.csv(submit, file = "condinfforest_m2.csv", row.names = FALSE)
```


```{r famidcleanup}
# AK - families where Freq & FamilyID don't match:
FamIDs <- data.frame(table(combi$FamilyID))
FamIndex <- (sub("[A-Z][a-z]+", "\\1", famIDs$Var1) != as.character(famIDs$Freq))
FamilyOdd <- famIDs[FamIndex,] #193 out of 928 FamilyIDs are ODD
# Cleaning up
OddRows <- combi[which(combi$FamilyID %in% FamilyOdd$Var1[1]),]
for (i in 2:nrow(FamilyOdd)) {
    OddRows <- rbind(OddRows, combi[which(combi$FamilyID %in% FamilyOdd$Var1[i]),])
}
#SEE manual clean up doc
combi$FamilyIDCleanUp <- as.character(combi$FamilyID)
combi$FamilyIDCleanUp2 <- as.character(combi$FamilyIDCleanUp2)
combi$FamilySizeCleanUp <- combi$FamilySize
#run the cleaning stuff
combi$FamilyIDCleanUp <- as.factor(combi$FamilyIDCleanUp)
combi$FamilySizeCleanUp <- as.integer(combi$FamilySizeCleanUp)
combi$FamilyIDCleanUp2 <- combi$FamilyIDCleanUp
combi$FamilyIDCleanUp2[combi$FamilySizeCleanUp <= 2] <- 'Small'
combi$FamilySizeCleanUp <- as.integer(combi$FamilySizeCleanUp)
combi$FamilyIDCleanUp2 <- as.factor(combi$FamilyIDCleanUp2)

combi$FamilyID <- factor(combi$FamilyID)
#familytable <- aggregate(FamilyID ~ FamilyID + Name + FamilySize + PassengerId, data=combi, FUN=length) # how many are in a given Family ID
#familytable <- aggregate(FamilyID ~ FamilySize, data=combi, FUN=length) # how many are in a given Family ID
#Preping FamilyID for Random Forests (<32 factor levels)
combi$FamilyID2 <- combi$FamilyID
combi$FamilyID2 <- as.character(combi$FamilyID2)
combi$FamilyID2[combi$FamilySize <= 3] <- 'Small'
combi$FamilyID2 <- factor(combi$FamilyID2)
# FamilyID - single person + three persons - to comp w/random forests
table(combi$FamilySize)
table(combi$FamilyID)
combi$FamilyID3 <- as.character(combi$FamilyID)
combi$FamilyID3[combi$FamilySize == 1] <- 'Single'
combi$FamilyID3[combi$FamilySize == 2] <- 'Two'
combi$FamilyID3[combi$FamilySize == 3] <- 'Three'
combi$FamilyID3 <- factor(combi$FamilyID3)
table(combi$FamilyID3, combi$FamilySize)
table(combi$FamilyID3, combi$FamilySize, combi$Parch)
aggregate(FamilyID3 ~ SibSp + Parch, data=combi, FUN=function(x) {round(sum(x)/length(x), 3)})
```


```{r missingvalues}
## Imputing missing values
summary(combi) #Age, Embarked, Fare
#Age
Agefit <- rpart(Age ~ Pclass + Sex + SibSp + Parch + Fare + Embarked + Title + FamilySize,
                data=combi[!is.na(combi$Age),], method="anova")
combi$Age[is.na(combi$Age)] <- predict(Agefit, combi[is.na(combi$Age),])
#Embarked - 2 do not have values
which(combi$Embarked == '')
combi$Embarked[c(62,830)] = "S" #most common
#Fare - 1 does not have a value
which(is.na(combi$Fare))
combi$Fare[1044] <- median(combi$Fare, na.rm=TRUE)

# Split sets
train <- combi[1:891,]
test <- combi[892:1309,]

#### MODEL 1
#Random forest
fit <- randomForest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize +
                        FamilyID2, data=train, importance=TRUE, ntree=2000)
varImpPlot(fit) #importance of variables

Prediction <- predict(fit, test)
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "randomforest.csv", row.names = FALSE)


#### MODEL 2
# Forest of conditional inference trees - can handle large factors
fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID1,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m2b.csv", row.names = FALSE)


#### MODEL 3 - Adding "Single" as a category for FamilyID
## Compare to Random Forest
fit <- randomForest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize +
                        FamilyID3, data=train, importance=TRUE, ntree=2000)
varImpPlot(fit) #importance of variables
Prediction <- predict(fit, test)
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "randomforest_m3.csv", row.names = FALSE)

#### MODEL 4 - Using new FamilyID3 with Conditional Forests
fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID3,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m4.csv", row.names = FALSE)


#### MODEL 5 - 
# Families with infants?
# Male parent of the only child?
combi$SingleParent <- as.character(combi$SingleParent)
combi$SingleParent <- 'NotRemarkable'
#combi$SingleParent[(combi$Parch == 1) & (combi$Title %in% c('Mrs', 'Mr','Col', 'Dr', 'Lady', 'Rev','Sir'))] <- 'SingleParent'
combi$SingleParent[(combi$Parch == 1) & (combi$Title %in% c('Mr','Col', 'Dr', 'Rev','Sir'))] <- 'SingleFather'
combi$SingleParent[(combi$Parch == 1) & (combi$Title %in% c('Miss', 'Master'))] <- 'OnlyChild'
combi$SingleParent[(combi$Parch == 1) & (combi$Title %in% c('Miss', 'Master')) & (combi$Age <= 1)] <- 'Infant'
combi$SingleParent <- factor(combi$SingleParent)

combi$ChildParent <- as.character(combi$ChildParent)
combi$ChildParent <- 'NotRemarkable'
combi$ChildParent[(combi$Parch == 1) & (combi$SibSp == 0) & (combi$Title %in% c('Mr','Col', 'Dr', 'Rev','Sir'))] <- 'SingleParentMale'
#combi$ChildParent[(combi$Parch == 1) & (combi$Title %in% c('Mrs', 'Lady'))] <- 'SingleParentFemale'
#combi$ChildParent[(combi$Parch == 1) & (combi$Title %in% c('Miss', 'Master')) & (combi$Age <= 16)] <- 'OnlyChildUnder16'
#combi$ChildParent[(combi$Parch == 1) & (combi$Title %in% c('Miss', 'Master')) & (combi$Age <= 12)] <- 'OnlyChildUnder12'
combi$ChildParent[(combi$Parch == 1) & (combi$Title %in% c('Miss', 'Master')) & (combi$Age < 6)] <- 'OnlyChildUnder6'
#combi$ChildParent[(combi$Parch == 1) & (combi$Title %in% c('Miss', 'Master')) & (combi$Age <= 1)] <- 'Infant'
combi$ChildParent[(combi$Title %in% c('Miss', 'Master')) & (combi$Age <= 1)] <- 'Infant'
combi$ChildParent <- as.factor(combi$ChildParent)

#more options - parent of infant
infants <- subset(combi, combi$Age <= 1) #infants
infants <- subset(combi, ((combi$Age <= 1) & (combi$Parch == 1))) #infants with single parent
parentIDs <- as.character(infants$FamilyID) # infants with single parent
smallchildren <- subset(combi, ((combi$Age > 1) & (combi$Age <= 3) & (combi$Parch == 1))) #with single parent
parentIDs <- as.character(c(infants$FamilyID, smallchildren$FamilyID))
parentIDsFreq <- data.frame(table(parentIDs))
parentIDs1 <- parentIDs[parentIDsFreq$Freq ==1]
parentIDs2 <- parentIDs[parentIDsFreq$Freq ==2]
combi$ChildParent <- as.character(combi$ChildParent)
combi$ChildParent[(combi$FamilyID %in% parentIDs1) & (combi$Title %in% c('Mrs', 'Mr','Col', 'Dr', 'Lady', 'Rev','Sir'))] <- 'SingleParentOfSmChild'
combi$ChildParent[(combi$FamilyID %in% parentIDs2) & (combi$Title %in% c('Mrs', 'Mr','Col', 'Dr', 'Lady', 'Rev','Sir'))] <- 'TwoParentsOfSmChild'
#combi$ChildParent[(combi$FamilyID %in% parentIDs) & (combi$Title %in% c('Mrs', 'Mr','Col', 'Dr', 'Lady', 'Rev','Sir'))] <- 'ParentOfInfant'
combi$ChildParent[(combi$Title %in% c('Miss', 'Master')) & (combi$Age <= 1)] <- 'Infant'
#combi$ChildParent[(combi$Parch == 1) & (combi$Title %in% c('Mrs', 'Lady'))] <- 'SingleParentFemale'

combi$ChildParent <- factor(combi$ChildParent)
combi$FamilyIDCleanUp <- factor(combi$FamilyIDCleanUp)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID3 + singleParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m5.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID1 + singleParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m6.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID3 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m7.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID3 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m8.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID3 + SingleParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m9.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID3 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m10.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID1 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m11.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID1 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m12.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySize + FamilyID1 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m13.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySizeCleanUp + FamilyIDCleanUp2,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m14.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title + FamilySizeCleanUp + FamilyIDCleanUp2 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m15.csv", row.names = FALSE)

fit <- cforest(as.factor(Survived) ~ Pclass + Sex + Age + Fare + Title + 
                   FamilySizeCleanUp + FamilyIDCleanUp2 + ChildParent,
               data = train, controls=cforest_unbiased(ntree=2000, mtry=3))
Prediction <- predict(fit, test, OOB=TRUE, type = "response")
submit <- data.frame(PassengerId = test$PassengerId, Survived = Prediction)
write.csv(submit, file = "conditionalforest_m16.csv", row.names = FALSE)

####
write.csv(combi, file = "combi.csv", row.names = FALSE)
```

You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot(cars)
```

